Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-18 vertical-thermal-review.Rmd
citation update

Leaf hydraulic, stomatal conductance, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light--– typically tracking photosynthetic capacity (Wong et al. 1979)-– and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, Yoder et al. 1994, Koch et al. 2004, Sillet et al. 2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations. These constraints tend to educe leaf water potential making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
Stomatal depression duces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.

>>>>

Leaf hydraulic, stomatal conductance, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light--– typically tracking photosynthetic capacity (Wong et al. 1979)-– and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, @yoderEvidenceReducedPhotosynthetic1994; @kochLimitsTreeHeight2004; @sillettIncreasingWoodProduction2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations. These constraints tend to educe leaf water potential making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
Stomatal depression duces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021]. This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
